\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-05-16 Paper/chapters/implementation-strategies.tex
camellia_128_sliced working

% TODO change shr/shl implementation to only use vector registers
One caveat of this approach is the fact that NEON registers cannot be shifted
in their entirety due to the fact bits are not able to cross lanes. This leads
to the problem of being able to shift at most two lanes of 64 bits at once. We
thus need to implement the \texttt{shr(V,n)} and \texttt{shl(V,n)} operations
on our own. This can be done by first extracting the 64-bit lanes $a,b$ out of
$V=b||a$, shifting the lanes individually and finally shifting and ORing the
crossing bits back into the other lane.

\begin{alignat*}{2}
    &\text{shl}(V,n): \\
    &\qquad a,b&&=V[0],V[1] \\
    &\qquad c&&=(a\gg (64-n)) \\
    &\qquad V[0]&&=(a\ll n) \\
    &\qquad V[1]&&=(b\ll n)\lor c
\end{alignat*}

@@

\begin{align*}
    &\text{SWAPMOVE}(V_0,V_1,0x5555\dots 55,1) &\text{SWAPMOVE}(V_4,V_5,0x5555\dots 55,1) \\
    &\text{SWAPMOVE}(V_2,V_3,0x5555\dots 55,1) &\text{SWAPMOVE}(V_6,V_7,0x5555\dots 55,1) \\
    &\text{SWAPMOVE}(V_0,V_2,0x3333\dots 33,2) &\text{SWAPMOVE}(V_4,V_6,0x3333\dots 33,2) \\
    &\text{SWAPMOVE}(V_1,V_3,0x3333\dots 33,2) &\text{SWAPMOVE}(V_5,V_7,0x3333\dots 33,2) \\
    &\text{SWAPMOVE}(V_0,V_4,0x0f0f\dots 0f,4) &\text{SWAPMOVE}(V_1,V_5,0x0f0f\dots 0f,4) \\
    &\text{SWAPMOVE}(V_2,V_6,0x0f0f\dots 0f,4) &\text{SWAPMOVE}(V_3,V_7,0x0f0f\dots 0f,4) \\
\end{align*}

@@

\subsubsection{Packing}

We encrypt 16 plaintext blocks in parallel so we can fill 16 vector registers
with register $i$ containing all $i$-th bytes. This lends itself well to a NEON
implementation since S-boxes as well as the FL layer, Feistel round and
packing/unpacking functions can be implemented efficiently.

@@

with affine transformations $h,f$ and the multiplicative inversion function $g$
in the composite field GF$((2^4)^2)$. Because Galois fields of equal size are
isomorphic, there are affine isomorphisms $\delta,\delta^{-1}$ between
GF($2^8$) and GF($(2^4)^2$) respectively\cite{satohmorioka:2003}:

@@

>>>>

% TODO change shr/shl implementation to only use vector registers
One caveat of this approach is the fact that NEON registers cannot be shifted
in their entirety due to the fact bits are not able to cross lanes. This leads
to the problem of being able to shift at most two lanes of 64 bits at once. We
thus need to implement the \texttt{shr(V,n)} and \texttt{shl(V,n)} operations
on our own. This can be done by extracting and shifting the overflow $ov$ out
of $V=b||a$, shifting the lanes individually and finally ORing the overflow
bits to the corresponding vector element.

\begin{alignat*}{2}
    &\text{shl}(V,n): \\
    &\qquad ov&&=V[0]\land(\lnot(\texttt{0xffffffffffffffff}\gg_n))\gg_{63-n} \\
    &\qquad V[0]&&=V[0]\gg_n \\
    &\qquad V[1]&&=(V[1]\gg_n)\lor ov
\end{alignat*}

@@

\begin{align*}
    &\text{SWAPMOVE}(V_0,V_1,0x5555\dots 55,1) &\text{SWAPMOVE}(V_4,V_5,0x5555\dots 55,1) \\
    &\text{SWAPMOVE}(V_2,V_3,0x5555\dots 55,1) &\text{SWAPMOVE}(V_6,V_7,0x5555\dots 55,1) \\
    &\text{SWAPMOVE}(V_0,V_2,0x3333\dots 33,2) &\text{SWAPMOVE}(V_4,V_6,0x3333\dots 33,2) \\
    &\text{SWAPMOVE}(V_1,V_3,0x3333\dots 33,2) &\text{SWAPMOVE}(V_5,V_7,0x3333\dots 33,2) \\
    &\text{SWAPMOVE}(V_0,V_4,0x0f0f\dots 0f,4) &\text{SWAPMOVE}(V_1,V_5,0x0f0f\dots 0f,4) \\
    &\text{SWAPMOVE}(V_2,V_6,0x0f0f\dots 0f,4) &\text{SWAPMOVE}(V_3,V_7,0x0f0f\dots 0f,4)
\end{align*}

@@

Because Camellia is a byte-oriented block cipher, we can pursue a similar
strategy as for GIFT: find a convenient data packing format and a way to apply
necessary operations in parallel.

We choose to encrypt 16 plaintext blocks at once so we can fill 16 vector
registers with register $V_i$ containing all $i$-th bytes. This lends itself well
to a NEON implementation since S-boxes as well as the FL layer, Feistel round
and packing/unpacking functions can be implemented efficiently.

\subsubsection{Packing and unpacking}

Packing and unpacking of data can be done efficiently by use of 32 \texttt{TBL}
instructions and is summarized in Figure \ref{fig:byteslicing}. After
packing, every $i$-th register will contain all $i$-th bytes of the 16 input
blocks. Unpacking is done in a similar way with different permutation tables.

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/byteslicing.pdf}
    \caption{Camellia byte-sliced packing function with 32 \texttt{TBL} operations.
    Curly bracets represent a 4-element vector array being used as source registers.}
    \label{fig:byteslicing}
\end{figure}

@@

with affine transformations $h,f$ and the multiplicative inversion function $g$
in the composite field GF$((2^4)^2)$. Because Galois fields of equal size are
isomorphic, there exist affine isomorphisms $\delta,\delta^{-1}$ between
GF($2^8$) and GF($(2^4)^2$) respectively\cite{satohmorioka:2003}:

@@

\subsubsection{Permutation}

Since all bytes of the same position are collected in the same register,
applying the permutation can be realized in 16 XOR operations as described in
the Camellia specification with the difference that we choose to not swap the
lower and higher 4 bytes, but rather compensate for this by choosing the
correct bytes when XORing with the other half later in the Feistel round.

% TODO add graphic here

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/camelliaround.pdf}
    \caption{Camellia byte-sliced Feistel round function}
\end{figure}

\subsubsection{FL layer}

The two functions $FL,FL^{-1}$ can be implemented in a straightforward manner
once 1-bit left rotation has been defined for the bytesliced representation.
This can be facilitated by extracting the highest bit of each byte, shifting it
all the way to the right and storing it as overflow $o_i$. $o_i$ is then added
to the left-shifted value of byte $(i + 4) (\bmod 4)$ for the final result.

% TODO add graphic here
